---
output:
  md_document:
    variant: markdown_github
---

```{r, knitr-options, echo = FALSE}
# rmarkdown::render("README.Rmd")
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "README/",
  cache = FALSE)
```

# multiline

multiline is an R package for reading data from multiline fixed-width-formatted
(FWF) files. This format is like that of typical FWF files, except that data for
a given observation wraps after some number of columns.

## Background

Consider the following multiline FWF (MFWF) data. As with FWF data, parsing
requires the column positions of each field. But furthermore, we need the line
position of each field.

```
123456789
789
123456789
789
```

If `field1` appears in columns 1-9 of line 1 and `field2` is in columns 1-3 of
line 2, then the data give two observations with identical values for each
field.

Thus another way to see MFWF data is as a "long" table (i.e., with observations
spanning rows) whose unique identifiers are implicit. The updated example below
shows these observation identifiers in parentheses and line numbers in brackets.

```
123456789  (1) [1]
789        (1) [2]
123456789  (2) [1]
789        (2) [2]
```

Because line numbers in MFWF data are inferred (just as column numbers in FWF
data), lines must appear whether or not they contain data. The updated example
below shows how the data would look if `field2` were missing for observation 1.
Again, the parentheses show observation identifiers and brackets show line numbers.

```
123456789  (1) [1]
           (1) [2]
123456789  (2) [1]
789        (2) [2]
```


## Usage

We can represent MFWF field positions in a list that contains for each line a
table with its fields' names and start and end positions. The `fwf_positions`
function from `readr` makes this makes this easy. For example, if `field1` is in
line 1, columns 1-9, and `field2` is in line 2, columns 1-3, then:

```{r}
positions = list(
  readr::fwf_positions(start = 1, end = 9, col_names = 'field1'),
  readr::fwf_positions(start = 1, end = 3, col_names = 'field2'))
positions
```

The first table in the list "positions" gives the column positions of fields 
in the first line of the data, and the second table in the list "positions" 
gives the column positions of fields in the second line of the data. 

Given the data:

```{r}
d <- "123456789\n789\n123456789\n789"
```

`read_multiline()` returns a tidy table with observations in rows and fields in
columns.
Note that `read_multiline()` requires that the number of items in the list of 
positions exactly matches the number of lines in the MFWF. 

```{r}
tidy <- read_multiline(d, lines=2, positions)
tidy
```


## Installation

Install from GitHub with devtools:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("jamesdunham/multiline")
```

